A free/open-source Kazakh-Tatar machine translation system

### 2013-04-17 paper/paper.tex
a bunch of copy-editing

%Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal] (сен, сендер, сіз, сіздер), whereas Tatar collapse all pronouns except the [-plural, -formal] син into one pronoun (сез).  This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}  
\begin{table}[htbp]
%\begin{subtables}
	%\begin{subtable}[l]
	%\subtable[Kazakh 2nd person pronouns]{
	%\begin{subtable}
	\subfloat[Kazakh 2nd person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & -pl & +pl \\\midrule
			-formal & сен & сендер \\
			+formal & сіз & сіздер \\\bottomrule
		\end{tabular}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	%\begin{subtable}
	%\begin{subtable}
	%}\subtable[Tatar 2nd person pronouns]{
	}\vspace{1.5ex}
	\subfloat[Tatar 2nd person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & -pl & +pl \\\midrule
			-formal & син & сез \\
			+formal & сез & сез \\\bottomrule
		\end{tabular}%}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	}
	\caption{The 2nd person pronoun systems of Kazakh and Tatar}
	\label{tab:2pprons}
\end{table}
%\end{subtables}
This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns are only used for emphasis and clarification.  However, this difference between Tatar and Kazakh in the second person system runs much deeper than just the pronoun system.  Since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (even when there is no overt pronoun), the Kazakh and Tatar systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes which have a one-to-one correspondence with the pronouns in each language.

> Past tense system

>>>>

%Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal] (сен, сендер, сіз, сіздер), whereas Tatar collapse all pronouns except the [-plural, -formal] син into one pronoun (сез).  This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
Another example of a far-reaching morphological difference between Tatar and Kazakh is the presence of a four-way distinction in Kazakh's 2nd person system (both pronouns and agreement suffixes), where Tatar only has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}.
\begin{table}[htbp]
%\begin{subtables}
	%\begin{subtable}[l]
	%\subtable[Kazakh 2nd person pronouns]{
	%\begin{subtable}
	\subfloat[Kazakh 2nd person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & -pl & +pl \\\midrule
			-formal & сен & сендер \\
			+formal & сіз & сіздер \\\bottomrule
		\end{tabular}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	%\begin{subtable}
	%\begin{subtable}
	%}\subtable[Tatar 2nd person pronouns]{
	}\vspace{1.5ex}
	\subfloat[Tatar 2nd person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & -pl & +pl \\\midrule
			-formal & син & сез \\
			+formal & сез & сез \\\bottomrule
		\end{tabular}%}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	}
	\caption{The 2nd person pronoun systems of Kazakh and Tatar}
	\label{tab:2pprons}
\end{table}
%\end{subtables}

This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns are only used for emphasis and clarification.  However, this difference between Tatar and Kazakh in the second person system runs much deeper than just the pronoun system.  Since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (even when there is no overt pronoun, in either situation), the Kazakh and Tatar systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes which have a one-to-one correspondence with the pronouns in each language, resulting in several systems of suffixes in each language that have the same set of distinctions as in the 2nd person pronoun systems.

The past tense systems of Kazakh and Tatar have a many-to-many correspondence.  At a basic level, in the past tense, Kazakh differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.  As an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness past tense morpheme (\texttt{-GAн-}) while Kazakh has a recent non-eyewitness past (\texttt{-Iп-}) and a distant non-eyewitness past (\texttt{-GAн екен-}).  On the other hand, these two non-eyewitness past forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.

Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are different.  Specifically, in both languages, the person agreement takes the form of a person copula suffix, although in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in Tatar this suffix intervenes between the two pieces of the ``compound'' tense morpheme (e.g., баргансың икән ``I guess you went'').

Another morphotactic difference between Kazakh and Tatar is found with the negative forms of the cognate \texttt{-GAн-} past tenses.  In Kazakh, the negative form of the non-recent reliable-information past tense is \texttt{-GAн емес-}, whereas in Tatar, the negative form of the non-eyewitness past tense is \texttt{-мAGAн-}.

%Morphotactics of -GAн екен/икән
